Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
global props

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the SFD
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, 99\% of the stars in our sample have at least 8 visibility periods,
while the sample median and root-variance in number of visibility periods are 14
and 3, respectively, guaranteeing a robust estimate of proper motions.
The variance in the number density of faint stars originating from the \gaia\
scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
\mu_{\phi_2, \odot} < 4~\masyr$ (i.e. above the stream selection box), far
smaller than the density variations in the stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet begun to phase-mix in multiple
dimensions.

@@

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
During simulation, star particles are released from the Lagrange points of the
cluster with a spread in position and velocity set by the mass of the satellite
and the location within the host galaxy, with some tunable scale parameters
(using the parametrization and adopted scale parameters of \cite{Fardal:2015}).
This method can accurately reproduce the mean track of the resulting stream, but
the density of stars along the stream depends in detail on the mass-loss history
and internal kinematics of the progenitor system.
Here we assume a constant mass-loss rate, and release stream particles uniformly
in time from each Lagrange point; these choices match results from more detailed
numerical experiments of globular cluster disruption
\citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a disk (\citealt{Miyamoto:1975}), a bulge
(\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to closely match the disk model found in
\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.
We use L-BFGS-B optimization as implemented in the \package{scipy}
\texttt{Python} package to maximize the likelihood of the orbit given sky track
and proper motions from this work with mean distance and giant star radial
velocities from \cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison model of the stream properties.

@@

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We integrate backwards for $4~\textrm{Gyr}$ using the orbital parameters at this
location using the Leapfrog integration scheme with a timestep of
$0.5~\textrm{Myr}$, then forward-generate the model stream using the method
discussed above.
Stream particles are release at each timestep of the integrated progenitor
orbit.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.

>>>>

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the SFD
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
periods (median and root-variance of 14 and 3, respectively), guaranteeing a
robust estimate of proper motions.
The variance in the number density of faint stars originating from the \gaia\
scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
\mu_{\phi_2, \odot} < 4~\masyr$ (i.e. above the stream selection box), far
smaller than the density variations in the stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material that has not yet phase-mixed in multiple dimensions.

@@

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
We use the ``particle-spray'' stream generation method (\citealt{Fardal:2015})
and assume a uniform mass-loss history to generate the model stream, which
choices matches results from more detailed ($N$-body) numerical experiments of
globular cluster disruption \citep[e.g.,][]{Kupper:2012}.
% During simulation, star particles are released from the Lagrange points of the
% cluster with a spread in position and velocity set by the mass of the
% satellite and the location within the host galaxy, with some tunable scale
% parameters (using the parametrization and adopted scale parameters of
% \cite{Fardal:2015}).  This method can accurately reproduce the mean track of
% the resulting stream, but the density of stars along the stream depends in
% detail on the mass-loss history and internal kinematics of the progenitor
% system.  Here we assume a constant mass-loss rate, and release stream
% particles uniformly in time from each Lagrange point; these choices match
% results from more detailed numerical experiments of globular cluster
% disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a disk (\citealt{Miyamoto:1975}), a bulge
(\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to match the disk model found in
\citet{Bovy:2015}; other parameters are set to produce an approximately flat
disk rotation curve between with a circular velocity at the solar circle,
$v_{\textrm{circ}, \odot} \approx 230~\kms$.
We use L-BFGS-B optimization as implemented in the \package{scipy}
\texttt{Python} package to maximize the likelihood of the orbit given sky track
and proper motions from this work with mean distance and giant star radial
velocities from \cite{Koposov:2010}.
We do not vary the parameters of the Milky Way model in the fit because we are
only interested in producing a comparison model of the stream properties.

@@

To initialize the model stream generation, we assume that the present day (i.e.
final) position of the progenitor system is at $\phi_1 = -13.5^\circ$, the peak
of the observed surface density profile.
We numerically integrate the backwards from this location for $4~\textrm{Gyr}$
then forward-generate the model stream using the method discussed above, and
release stream particles at each timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until present day, where we assume $M = 0~\msun$
(i.e. fully disrupted).
Stream properties computed from this resulting model stream are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange) lines.
The model stream qualitatively matches the stream track in all phase-space
dimensions, but has a much smoother surface density profile along the extent of
the stream.
We therefore don't expect that the observed surface density variations could be
caused by, e.g., epicyclic over-densities (\citealt{Kupper:2012}).
